Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-13 ind_files/01.5_math_background.tex
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

\subsection{Birth-death processes}
Mathematically, a birth-death process is a so-called `continuous-time Markov chain' in which only transitions between local states are allowed. In other words, a birth-death process is a stochastic process unfolding in continuous time such that
\begin{itemize}
	\item The process is `Markov', meaning that the future is statistically independent of the past given the present. In more mathematical terms, if the value of the stochastic process at time $t$ is given by $X_t$, $\mathbb{P}(\cdot | E)$ denotes probability conditioned on $E$, and $u < s \leq t$, then
	\begin{equation*}
	\mathbb{P}(X_t | X_s, X_u) = \mathbb{P}(X_t | X_s)
	\end{equation*}
	\item Direct transitions must be `local'. Mathematicians usually reserve the phrase `birth-death process' to processes that take values in the non-negative integers $\{0,1,2,3,4,\ldots\}$. In this case, only direct transitions from $n$ to $n \pm 1$ are allowed to occur. Biologically, this is saying that we observe the population on a fine enough timescale that the probability of two or more births/deaths occurring at the exact same time is very low and we can disallow it entirely in our models. The conditions for higher dimensional birth-death processes look similar.
\end{itemize}
Since these processes unfold in continuous time, they are characterized not by transition probabilities but by transition \emph{rates}, which can be thought of as the probability of transition `per unit time'. The quantity of interest is usually the probability of being in a particular state at a given point in time. The entire birth-death process can be described in terms of such a quantity, through a so-called `Master equation'. The master equation is a partial differential equation (PDE) for the probability of being in a given state at a given time, However, in all but the simplest cases, we can't actually solve this PDE, because it is simply too hard. The primary source of difficulty is non-linearity in the transition rates and the fact that transitions occur in discrete, discontinuous `jumps'. It is much easier to describe and analyze systems by using tools from stochastic calculus and partial differential equations, as we describe below.

\subsection{SDEs and the Fokker-Planck equation}\label{intro_SDE}
Stochastic systems which change continuously (in the state space) can be described in terms of a `stochastic differential equation' (SDE), which here is interchangeable with the phrase `It\^o process'. An SDE for a stochastic process $\{X_t\}_{t \geq 0}$ is an equation of the form
\begin{equation}
\label{ito_SDE_integral}
X_t = \int\limits_{0}^{t} F(s,X_s)ds + \int\limits_{0}^{t} G(s,X_s)dB_s
\end{equation}
where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so we assume going further that all our SDEs are always well-defined and admit solutions.} In the physics literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, we will not use this terminology here due to  potential confusion with genetic drift (which actually corresponds to the `diffusion' in the physics terminology). $B_t$ is the so-called `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{B_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $B_t$ as recording the position of a small pollen grain at time $t$, then $B_t$ can be formally thought of as a process that has the following properties:
\begin{itemize}
	\item It starts at the origin, \emph{i.e} $B_0 = 0$. This is a harmless assumption made for convenience and amounts to a choice of coordinate system.
	\item It moves continuously, without sudden jumps across regions of space, \emph{i.e} the map $t \to B_t$ is continuous. This simply says that our pollen grain moves short distances in short intervals of time.
	\item The future movement is independent of past history. That is, given times $u < s < t$, the displacement $B_t - B_s$ is independent of the past position $B_u$.
	\item The movement is directionless and random, and displacement is normally distributed. More precisely, given two times $s < t$, the displacement $B_t - B_s$ follows a normal distribution with a mean of $0$ (this is the `directionless' part) and a variance of $t-s$ (this is the `random' part).
\end{itemize}
It can then be shown that since the motion is equally likely to be in any direction, the expected position at any point of time is the same as the initial position, \emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$.

The second integral in equation \eqref{ito_SDE_integral} is It\^o's `stochastic integral', and is to be interpreted in the following sense: Fix a time $T > 0$. For any $n\geq2$, let $\Pi_n = \{t_1,t_2,\ldots,t_n\}$ be a partition of the interval $[0,T]$. In other words, $\Pi_n$ divides $[0,T]$ into $n$ slices of the form $[t_i,t_{i+1}]$ such that $0 = t_0 < t_1 < t_2 < \ldots < t_n = T$. Then, the (It\^o) stochastic integral from $0$ to $T$ can be thought of as:
\begin{equation*}
\int\limits_{0}^{T} G(s,X_s)dB_s \coloneqq \lim_{n \to \infty} \sum\limits_{t_i \in \Pi_n}G(t_i,X_{t_i})(B_{t_{i+1}}-B_{t_i})  
\end{equation*}
That is to say, it is obtained by making successively finer partitions of the form $[t_i,t_{i+1}]$, and then computing the `area of the rectangle' formed with $B_{t_{i+1}}-B_{t_i}$ and $G(t_i,X_{t_i})$ as sides. This should look similar to the classic Riemann integral, with the uniform width $t_{i+1}-t_i$ of the Riemann integral replaced by a random width corresponding to the random displacement of a Brownian particle during the uniform time interval $[t_i,t_{i+1}]$.

@@

dX_t = F(t,X_t)dt + G(t,X_t)dB_t

@@

where $\eta(t)$ is supposed to be `Gaussian white noise', defined indirectly such that $\int_0^{t}G(s,x)\eta(s)ds$ behaves identically to $\int_0^{t}G(s,X_s)dB_s$. However, it is important to remember that these are both purely formal expressions - Equation \eqref{ito_SDE_diff} is meaningless on its own and is really just shorthand for equation \eqref{ito_SDE_integral}, which is well-defined as explained above; Equation \eqref{ito_langevin} is even worse, because the Brownian motion is known to be non-differentiable, and as such, $\eta(t)$ cannot really exist - Both equations are thus to be interpreted as shorthand for equation \eqref{ito_SDE_integral}, which formally `makes sense'. SDEs are convenient because they satisfy several `nice' analytical properties. For example, using the fact that the Brownian motion has no expected change in value (\emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$), it can be shown\footnote{We can actually prove something stronger: We can show under rather mild regularity assumptions on $X_t$ and $G(t,x)$ that the stochastic integral is a continuous square-integrable martingale starting at the origin - This means that the map $t \to \int_0^{t}G(s,X_s)dB_s$ is continuous, starts at the origin, and always has an expectation value of $0$.} that the stochastic integral also has an expectation value of $0$ for all $t$, \emph{i.e}:

@@

\mathbb{E}\left[\int\limits_{0}^{t} G(s,X_s)dB_s \bigg{|} X_0\right] = 0

@@

		& $\mathbf{dt}$ & $\mathbf{dB_t}$ \\

@@

		$\mathbf{dB_t}$ & $0$  & $dt$ \\

@@

Naively, we may expect the same logic to still hold true for one-dimensional It\^o processes of the form \eqref{ito_SDE_diff} with $gdt$ simply being replaced by $GdB_t$ on the RHS. However, this does not work. The correct relation is instead given by 

@@

dh(X_t) = h'(X_t)F(X_t)dt + h'(X_t)G(X_t)dB_t  + \frac{h''(X_t)}{2}G^2(X_t)dt 

@@

\subsection{Density-dependence and the intuition for system-size expansions in ecology}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations in ecology, at an individual level, births and deaths of individuals are affected by local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In ecology, this parameter will be some fundamental limit on resources, such as habitat size or carrying capacity. In physics and chemistry, it is usually the total volume of a container in which physical or chemical reactions take place. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'\citep{ewens_mathematical_2004} or `continuum limit'~\citep{czuppon_understanding_2021}, and has been heavily used by Kimura~\citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed~\citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter (\emph{i.e.} is closer to a Kramers-Moyal expansion than a Van Kampen expansion).

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. Since all individuals are identical, we can only really track the population size through time. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', we mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate $b(n)$ of a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
\\

Exactly analogously, we also assume we can define a \emph{death rate} $d(n)$ of a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} has just occurred, then, the probability that the event that occurred is a birth is
\begin{equation*}
\mathbb{P[\textrm{ birth } | \textrm{ something happened }]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event was instead a death is given by
\begin{equation*}
\mathbb{P[\textrm{ death } | \textrm{ something happened }]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\\

@@

\subsection{The system-size expansion}
The system-size expansion arises from noting that in many systems, the interactions are governed not by population size, but by population \emph{density}. However, the population jumps themselves are discretized at the scale of the individual, which becomes negligibly small if we have a large population density. Thus, we assume that there exists a system-size parameter $K > 0$ such that the discrete jumps between states happen in units of $1/K$, and we make the substitutions
\begin{align*}
x &= \frac{n}{K}\\
b_K(x) &= \frac{1}{K}b(n)\\
d_K(x) &= \frac{1}{K}d(n)
\end{align*}
As $K$ grows very large, the discontinuous jumps in $n$ thus appear like `continuous' transitions in our new variable $x$, which can be thought of as the `density' of organisms. A system-size parameter $K$ often naturally emerges in ecological systems through resource-limiting factors such as habitat size or carrying capacity. Under these substitutions, equation \eqref{1D_M_eqn} becomes
\begin{equation}
\label{1D_M_eqn_density}
\frac{\partial P}{\partial t}(x,t) = (\Delta^{-}-1)Kb_K(x)P(x,t) + (\Delta^{+}-1)Kd_K(x)P(x,t)
\end{equation}
where we now have the new step operators
\begin{equation}
\label{1D_step_operators_density}
\Delta^{\pm}f(x,t) = f\left(x\pm\frac{1}{K},t\right) 
\end{equation}
If $K$ is large, then we can now taylor-expand the action of these step operators as:
\begin{equation*}
\Delta^{\pm}f(x,t) = f\left(x\pm\frac{1}{K},t\right) = f(x,t) \pm \frac{1}{K}\frac{\partial f}{\partial x}(x,t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x^2}(x,t) + \mathcal{O}(K^{-3})
\end{equation*}
Substituting these expansions into \eqref{1D_M_eqn_density} and neglecting terms of $\mathcal{O}(K^{-3})$ and higher, we obtain
\begin{equation}
\label{1D_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{
	\frac{\partial P}{\partial t}(x,t) = -\frac{ \partial}{\partial x}\{A^{-}(x)P(x,t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x^2}\{A^{+}(x)P(x,t)\}
}
\end{equation}
where
\begin{equation*}
A^{\pm}(x) = b_K(x) \pm d_K(x)
\end{equation*}
Equation \eqref{1D_FPE} has the form of a so-called `Fokker-Planck equation', and corresponds to the SDE:
\begin{equation}
\label{1D_SDE}
dX_t = A^{-}(X_t)dt + \sqrt{\frac{A^{+}(X_t)}{K}}dB_t
\end{equation}
interpreted in the It\^{o} sense. Note that the deterministic component of this process depends on the difference between birth and death rates (a mechanistic measure of Malthusian fitness), whereas the stochastic part depends on their sum and scales inversely with $\sqrt{K}$.

>>>>

\subsection{Birth-death processes}
Mathematically, a birth-death process is a so-called `continuous-time Markov chain' in which only transitions between local states are allowed. In other words, a birth-death process is a stochastic process unfolding in continuous time such that
\begin{itemize}
	\item The process is `Markov', meaning that the future is statistically independent of the past given the present. In more mathematical terms, if the value of the stochastic process at time $t$ is given by $X_t$, $\mathbb{P}(\cdot | E)$ denotes probability conditioned on $E$, and $u < s \leq t$, then
	\begin{equation*}
	\mathbb{P}(X_t | X_s, X_u) = \mathbb{P}(X_t | X_s)
	\end{equation*}
	This equation is simply saying that if we have the information about the state of a process at time $s$, then we do not gain any more predictive power about the process at a future time $t$ if we have additional knowledge about the process from some past time $u < s$. A series of tosses of a fair coin is a simple example of a Markov process, since knowing whether a coin landed on heads during a previous toss does not change your predictive power about whether the coin will land on heads the next time you toss it.
	\item Direct transitions must be `local'. Mathematicians usually reserve the phrase `birth-death process' to processes that take values in the non-negative integers $\{0,1,2,3,4,\ldots\}$. In this case, only direct transitions from $n$ to $n \pm 1$ are allowed to occur. Biologically, this is saying that we observe the population on a fine enough timescale that the probability of two or more births/deaths occurring at the exact same time is very low and we can disallow it entirely in our models. The conditions for higher dimensional birth-death processes look similar.
\end{itemize}
Since these processes unfold in continuous time, they are characterized not by transition probabilities but by transition \emph{rates}, which can be thought of as the probability of transition `per unit time'. The quantity of interest is usually the probability of being in a particular state at a given point in time. The entire birth-death process can be described in terms of such a quantity, through a so-called `Master equation'. The master equation is a partial differential equation (PDE) for the probability of being in a given state at a given time, However, in all but the simplest cases, we can't actually solve this PDE, because it is simply too hard. The primary source of difficulty is non-linearity in the transition rates and the fact that transitions occur in discrete, discontinuous `jumps'. It is much easier to describe and analyze systems by using tools from stochastic calculus and partial differential equations, as we describe below.

\subsection{SDEs and the Fokker-Planck equation}\label{intro_SDE}
Stochastic systems which change continuously (in the state space) can be described in terms of a `stochastic differential equation' (SDE), which here is interchangeable with the phrase `It\^o process'. An SDE for a stochastic process $\{X_t\}_{t \geq 0}$ is an equation of the form
\begin{equation}
\label{ito_SDE_integral}
X_t = \int\limits_{0}^{t} F(s,X_s)ds + \int\limits_{0}^{t} G(s,X_s)dW_s
\end{equation}
where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so we assume going further that all our SDEs are always well-defined and admit solutions.} In the physics literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, we will not use this terminology here due to  potential confusion with genetic drift (which actually corresponds to the `diffusion' in the physics terminology). $W_t$ is the so-called `Wiener process' or `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{W_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $W_t$ as recording the position of a small pollen grain at time $t$, then $W_t$ can be formally thought of as a process that has the following properties:
\begin{itemize}
	\item It starts at the origin, \emph{i.e} $W_0 = 0$. This is a harmless assumption made for convenience and amounts to a choice of coordinate system.
	\item It moves continuously, without sudden jumps across regions of space, \emph{i.e} the map $t \to W_t$ is continuous. This simply says that our pollen grain moves short distances in short intervals of time.
	\item The future movement is independent of past history. That is, given times $u < s < t$, the displacement $W_t - W_s$ is independent of the past position $W_u$.
	\item The movement is directionless and random, and displacement is normally distributed. More precisely, given two times $s < t$, the displacement $W_t - W_s$ follows a normal distribution with a mean of $0$ (this is the `directionless' part) and a variance of $t-s$ (this is the `random' part).
\end{itemize}
It can then be shown that since the motion is equally likely to be in any direction, the expected position at any point of time is the same as the initial position, \emph{i.e} $\mathbb{E}[W_t | W_0] = W_0 = 0$.

The second integral in equation \eqref{ito_SDE_integral} is It\^o's `stochastic integral', and is to be interpreted in the following sense: Fix a time $T > 0$. For any $n\geq2$, let $\Pi_n = \{t_1,t_2,\ldots,t_n\}$ be a partition of the interval $[0,T]$. In other words, $\Pi_n$ divides $[0,T]$ into $n$ slices of the form $[t_i,t_{i+1}]$ such that $0 = t_0 < t_1 < t_2 < \ldots < t_n = T$. Then, the (It\^o) stochastic integral from $0$ to $T$ can be thought of as:
\begin{equation*}
\int\limits_{0}^{T} G(s,X_s)dW_s \coloneqq \lim_{n \to \infty} \sum\limits_{t_i \in \Pi_n}G(t_i,X_{t_i})(W_{t_{i+1}}-W_{t_i})  
\end{equation*}
That is to say, it is obtained by making successively finer partitions of the form $[t_i,t_{i+1}]$, and then computing the `area of the rectangle' formed with $W_{t_{i+1}}-W_{t_i}$ and $G(t_i,X_{t_i})$ as sides. This should look similar to the classic Riemann integral, with the uniform width $t_{i+1}-t_i$ of the Riemann integral replaced by a random width corresponding to the random displacement of a Brownian particle during the uniform time interval $[t_i,t_{i+1}]$.

@@

dX_t = F(t,X_t)dt + G(t,X_t)dW_t

@@

where $\eta(t)$ is supposed to be `Gaussian white noise', defined indirectly such that $\int_0^{t}G(s,x)\eta(s)ds$ behaves identically to $\int_0^{t}G(s,X_s)dW_s$. However, it is important to remember that these are both purely formal expressions - Equation \eqref{ito_SDE_diff} is meaningless on its own and is really just shorthand for equation \eqref{ito_SDE_integral}, which is well-defined as explained above; Equation \eqref{ito_langevin} is even worse, because the Brownian motion is known to be non-differentiable, and as such, $\eta(t)$ cannot really exist - Both equations are thus to be interpreted as shorthand for equation \eqref{ito_SDE_integral}, which formally `makes sense'. SDEs are convenient because they satisfy several `nice' analytical properties. For example, using the fact that the Brownian motion has no expected change in value (\emph{i.e} $\mathbb{E}[W_t | W_0] = W_0 = 0$), it can be shown\footnote{We can actually prove something stronger: We can show under rather mild regularity assumptions on $X_t$ and $G(t,x)$ that the stochastic integral is a continuous square-integrable martingale starting at the origin - This means that the map $t \to \int_0^{t}G(s,X_s)dW_s$ is continuous, starts at the origin, and always has an expectation value of $0$.} that the stochastic integral also has an expectation value of $0$ for all $t$, \emph{i.e}:

@@

\mathbb{E}\left[\int\limits_{0}^{t} G(s,X_s)dW_s \bigg{|} X_0\right] = 0

@@

		& $\mathbf{dt}$ & $\mathbf{dW_t}$ \\

@@

		$\mathbf{dW_t}$ & $0$  & $dt$ \\

@@

Naively, we may expect the same logic to still hold true for one-dimensional It\^o processes of the form \eqref{ito_SDE_diff} with $gdt$ simply being replaced by $GdW_t$ on the RHS. However, this does not work. The correct relation is instead given by 

@@

dh(X_t) = h'(X_t)F(X_t)dt + h'(X_t)G(X_t)dW_t  + \frac{h''(X_t)}{2}G^2(X_t)dt 

@@

\subsection{Density-dependence and the intuition for system-size expansions in ecology}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations in ecology, at an individual level, births and deaths of individuals are affected by local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In ecology, this parameter will be some fundamental limit on resources, such as habitat size or carrying capacity. In physics and chemistry, it is usually the total volume of a container in which physical or chemical reactions take place. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'\citep{ewens_mathematical_2004} or `continuum limit'~\citep{czuppon_understanding_2021}, and has been heavily used by Kimura~\citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed~\citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter.

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. Since all individuals are identical, we can only really track the population size through time. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', we mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate\footnote{Note that unlike usual ecology convention, this is \emph{not} a per-capita birth rate. The ecology per capita birth and death rates can be found by dividing my birth/death rates by the current population size $n(t)$} $b(n)$ of individuals a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
Exactly analogously, we can also define the \emph{death rate} $d(n)$ of individuals in a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} has just occurred, then, the probability that the event that occurred is a birth is
\begin{equation*}
\mathbb{P\big[\textrm{ birth } \big{|} \textrm{ something happened }\big]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event was instead a death is given by
\begin{equation*}
\mathbb{P\big[\textrm{ death } \big{|} \textrm{ something happened }\big]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\\

@@

\subsection{The system-size expansion}
The system-size expansion arises from noting that in many systems, the interactions are governed not by population size, but by population \emph{density}. However, the population jumps themselves are discretized at the scale of the individual, which becomes negligibly small if we have a large population density. Thus, we assume that there exists a system-size parameter $K > 0$ such that the discrete jumps between states happen in units of $1/K$, and we make the substitutions
\begin{align*}
x &= \frac{n}{K}\\
b_K(x) &= \frac{1}{K}b(n)\\
d_K(x) &= \frac{1}{K}d(n)
\end{align*}
As $K$ grows very large, the discontinuous jumps in $n$ thus appear like `continuous' transitions in our new variable $x$, which can be thought of as the `density' of organisms. A system-size parameter $K$ often naturally emerges in ecological systems through resource-limiting factors such as habitat size or carrying capacity. Under these substitutions, equation \eqref{1D_M_eqn} becomes
\begin{equation}
\label{1D_M_eqn_density}
\frac{\partial P}{\partial t}(x,t) = (\Delta^{-}-1)Kb_K(x)P(x,t) + (\Delta^{+}-1)Kd_K(x)P(x,t)
\end{equation}
where we now have the new step operators
\begin{equation}
\label{1D_step_operators_density}
\Delta^{\pm}f(x,t) = f\left(x\pm\frac{1}{K},t\right) 
\end{equation}
If $K$ is large, then we can now taylor-expand the action of these step operators as:
\begin{equation*}
\Delta^{\pm}f(x,t) = f\left(x\pm\frac{1}{K},t\right) = f(x,t) \pm \frac{1}{K}\frac{\partial f}{\partial x}(x,t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x^2}(x,t) + \mathcal{O}(K^{-3})
\end{equation*}
Substituting these expansions into \eqref{1D_M_eqn_density} and neglecting terms of $\mathcal{O}(K^{-3})$ and higher, we obtain
\begin{equation}
\label{1D_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{
	\frac{\partial P}{\partial t}(x,t) = -\frac{ \partial}{\partial x}\{A^{-}(x)P(x,t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x^2}\{A^{+}(x)P(x,t)\}
}
\end{equation}
where
\begin{equation*}
A^{\pm}(x) = b_K(x) \pm d_K(x)
\end{equation*}
Equation \eqref{1D_FPE} has the form of a so-called `Fokker-Planck equation', and corresponds to the SDE:
\begin{equation}
\label{1D_SDE}
dX_t = A^{-}(X_t)dt + \sqrt{\frac{A^{+}(X_t)}{K}}dW_t
\end{equation}
interpreted in the It\^{o} sense. Note that the deterministic component of this process depends on the difference between birth and death rates (a mechanistic measure of Malthusian fitness), whereas the stochastic part depends on their sum and scales inversely with $\sqrt{K}$.
